---
layout: topic
title: Introduction to R
author: Data Carpentry contributors
minutes: 45
---


```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ## Learning Objectives
>
> * Understand the concepts of vector and data types

------------

## Vectors and data types

```{r, echo=FALSE, purl=TRUE}
### Vectors and data types
```

A vector is the most common and basic data structure in R, and is pretty much
the workhorse of R. It's a group of values, mainly either numbers or
characters. You can assign this list of values to a variable, just like you
would for one item. For example we can create a vector of animal weights:

```{r, purl=FALSE}
weights <- c(50, 60, 65, 82)
weights
```

A vector can also contain characters:

```{r, purl=FALSE}
animals <- c("mouse", "rat", "dog")
animals
```

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r, purl=FALSE}
length(weights)
length(animals)
```

An important feature of a vector, is that all of the elements are the same type of data.
The function `class()` indicates the class (the type of element) of an object:

```{r, purl=FALSE}
class(weights)
class(animals)
```

The function `str()` provides an overview of the object and the elements it
contains. It is a really useful function when working with large and complex
objects:

```{r, purl=FALSE}
str(weights)
str(animals)
```

You can add elements to your vector by using the `c()` function:

```{r, purl=FALSE}
weights <- c(weights, 90) # adding at the end of the vector
weights <- c(30, weights) # adding at the beginning of the vector
weights
```

What happens here is that we take the original vector `weights`, and we are
adding another item first to the end of the other ones, and then another item at
the beginning. We can do this over and over again to grow a vector, or assemble
a dataset. As we program, this may be useful to add results that we are
collecting or calculating.

We just saw 2 of the 6 **atomic vector** types that R uses: `"character"` and
`"numeric"`. These are the basic building blocks that all R objects are built
from. The other 4 are:

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's an integer)
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
  `1+4i`) and that's all we're going to say about them
* `"raw"` that we won't discuss further

## Subsetting Vectors

If we want to extract individual values from within a vector, we can use the
square brackets `[]` with the position number inside the brackets:

```{r, purl=FALSE}
animals[2]
```

We can also extract multiple values using the combine `c()` function within the
square brackets:

```{r, purl=FALSE}
weights[c(1, 3)]
```

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`) and
factors (`factor`).

Next, we will use the "surveys" dataset to explore the
`data.frame` data structure, which is one of the most common types of R objects.

### Challenge

* **Question**: We’ve seen that atomic vectors can be of type character,
  numeric, integer, and logical. But what happens if we try to mix these types in
  a single vector?
* _Answer_: R implicitly converts them to all be the same type

* **Question**: Why do you think it happens?
* _Answer_: Vectors can be of only one data type. R tries to convert (=coerce)
  the content of this vector to find a "common denominator".

* **Question**: Can you draw a diagram that represents the hierarchy of the data
  types?
* _Answer_: `logical --> numeric --> character <-- logical`

```{r, echo=TRUE, eval=FALSE, purl=TRUE}
## We’ve seen that atomic vectors can be of type character, numeric, integer, and
## logical. But what happens if we try to mix these types in a single
## vector?

## What will happen in each of these examples? (hint: use `class()` to
## check the data type of your object)
num_char <- c(1, 2, 3, 'a')

num_logical <- c(1, 2, 3, TRUE)

char_logical <- c('a', 'b', 'c', TRUE)

tricky <- c(1, 2, 3, '4')

## Why do you think it happens?

## Can you draw a diagram that represents the hierarchy of the data
## types?
```
